When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-09-21 code/Submissionv3.Rmd
commit

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. In doing so, this paper develops a theory which argues that increasing political polarization in the United States impacts the military as a profession by straining the individual military officer's concept of loyalty in important ways. Drawing on an original data set of more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold professional standards as defined by historical military norms and standards. The paper concludes by arguing that the concept of "politicization of the military" --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the novel phenomenon of retired military officers using the concept of loyalty to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In many cities and communities across the nation, largely-peaceful protests were followed by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America by authoring opinion-editorial publications. It is a question worthy of examination for many reasons, including that at its root, at least three factors are involved, each of which may shape civil-military relations: military professionalism, political polarization, and institutional credibility of the military. The main argument of this paper is that a weakening of the concept of military professionalism can clearly be seen in the way retired military officers have made arguments in the public square over the past four decades, that the main variable that can explain this variation is increasing political polarization within American Society. Further, this paper argues that the military institution is particularly susceptible to increasing political polarization because of the way the concept of loyalty is used to make arguments that contain a clear partisan or political leaning. These developments reveal the extent to which increasing political polarization is synonymous with a fractured domestic public, and highlight the extraordinary challenge that the US military will llikely face to remain a  professional in an enduring polarized political climate.

This paper proceeds in four parts. First, the paper briefly reviews the concept of military professionalism in order to orient the reader to the central role this concept has played in the ethos of the modern American military. Second, the paper builds a theory that demonstrates how certain features of today's polarized political landscape challenge the concept of loyalty for the individual military professional. Third, the paper explores the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately shows that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen to both argue about topics that fall outside of the realm of traditional national security topics, and to express clear partisan preferences when making their arguments. The analysis also shows that this trend has worsened as the American political climate has become increasingly polarized. Fourth, the paper concludes by briefly exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting.

\singlespace
# Military Professionalism as a Foundational Trait
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is long and storied. As a concept, the idea of a professional military stems from the post-WW2 era, when the US military was locked in the Cold War struggle agains the Soviet Union. Perhaps the strongest and certainly one of the original intellectual founders of a professional military is the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the character of American liberal society. Huntington further argued that the key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

@@

Although a professional and non-partisan military is the theoretical goal of the institution, the practical record and results remain far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory, in one form or another, on more than one occasion. Nonetheless, academic theorists, military tradition, and contemporary military leaders have long argued for the importance of the military remaining grounded in professional conduct that is both non-partisan and apolitical. At its core, it is this type of professional conduct which ultimately postures the military to earn the trust and confidence of the American people.

To be clear, the concept of military professionalism does not impose a normative prohibition on the publication of an opinion piece, or the expression of other forms of public dialogue, by a military professional. In fact, one of the foundations of Huntington's concept of military professionalism is the fact that military leaders possess unique and specialized skills that civilians simply do not have. Moreover, multistar generals and admirals in the contemporary military possess in many cases upwards of 30 and sometimes 40 years of service, and it is entirely reasonable to conclude that these officers in particular have been privy to decisions and experiences occurring at or near the highest levels of government. In other words, their experiences are not confined to only tactical matters. They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service, and more.    
It is reasonable, therefore, to conclude that a professional opinion piece written by a military officer would take on three qualities. First, a professionally-written opinion piece would address a relevant military topic consistent with the military expertise of the author. In the US context in particular, where the military conducts a variety of roles and performs a variety of missions, this would include a wide swath of topics, while at the same time, certain topics would be considered out of the realm of relevant military expertise, such as partisan domestic politics. Second, from the perspective of both expertise and institutional credibility, a professional opinion piece would be based on the officer's personal experiences and be written from an individual's perspective so as not to infer that an argument is being made on behalf of the military. Third, a professionally-written opinion piece would be free of of partisan sentiment or expression, whether advocacy or denouncement. If one of these three qualities -- addressing a relevant military topic, written from the perspective of an individual officer's expertise, free of partisan sentiment -- fails to be upheld, there is an increasing likelihood that the officer's words can be conveyed and/or received as unprofessional.

If and when a military officer violates standards of military professionalism when conveying an opinion to the American public, his or her words can cause great harm, either to American society, the military as an instition, or a combination of both. To understand how and why this is the case, we will now turn our attention to understanding the impacts on military professionalism caused by two changes that have developed within American Society in the past 50 years. The first is a gradual increase in the American public's opinion of the military as an institution, and the second, more significant change is the steady increase in political polarization within American society.

@@

# When Words Become Weaponized - A Theory of Political Polarization and Military Professionalism

@@

What mechanisms might be at work through increasing political polarization that ultimately impact military professionalism? To answer this important question, I rely on the three major characteristics of the contemporary polarized political landscape previously identified by scholars of American Politics. These included, first, a sharp increase in the number and scope of issues that have been subsumed by partisan debate and disagreement (the phenomenon referred to as "conflict extension" [@layman_party_2002]); second, the increase in the strength of negative feelings held by one partisan group towards those of the other side (the rise of "negative partisanship" [@abramowitz_great_2018]); and third, an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. Each of these three features of the politically polarized landscape strain the concept of military professionalism in important ways, and taken together, comprise a significant challenge to the individual military professional and thus, to the professional nature of the military as a whole. These challenges are summarized in Table One, below.

@@

The first way in which the concept of military professionalism is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This occurs because as "conflict extension" [@layman_party_2002] occurs and an increasing number of issues are swallowed up by partisan battlegrounds, there is less politically neutral space for someone to stand on. When polarization levels in America were not as high as they are today, it was more possible for individuals to either stay out of politics, avoid partisan squabbles, and otherwise to stay politically neutral. Today's polarizing environment seems to have the opposite impact: polarized politics seem to thrust issues and the unfolding political fight at the individual citizen, even when the individual is not deliberately seeking confrontation. There is no theoretical reason to suspect that this mechanism would not impact military leaders. Military leaders are citizens of the Republic, and though the institutional ethos of the military profession is to remain apolitical, such an ideal is difficult to uphold if less and less neutral political space exists.

@@

However, it is also precisely on this point that the military institution is perhaps most vulnerable to increasing political polarization. I contend that the rise of identity politics can and will be harmful if a military professional's identity and the loyalty associated with that identity directly conflicts with his or her identity and loyalty as a professional soldier. As any soldier knows, and as civilians can reasonably understand,few traits are more sacred in the military than loyalty. Indeed, the virtue of loyalty is extolled in a host of codes and mantras that have played a central part of the American military since its inception, and loyalty is captured, either in word or in spirit, in a host of phrases and professional codes. In the middle of the the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918].

High levels of political polarization obfuscate the ideal of loyalty, however. As core issues that involve individual perceptions of moral right and wrong are at stake, individuals, to include military officers, are less likely to not become involved, attached, or otherwise invested in these issues. This is especially true when the language and appeal of concepts of loyalty are employed to justify one is or is not taking action over an issue. In a highly polarized political climate, there is increasing likelihood that two different people military officers may adopt completely different opinions regarding a contentious issue, each believing that the other is deeply and perhaps even morally wrong.

As a quick case in point, consider that in June 2020, following President Trump's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and politically striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@noauthor_letter_2020]. The letter's clear implication is that if you are serving in the Trump administration and if you graduated from the US Military Academy (i.e. Secretaries Esper and Pompeo), you have forsaken your oath of office, the US Constitution, and the values that West Point stands for.

@@

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table Two, below). The first is when political polarization and military credibility are both low. When these conditions hold, there is no anticipated drop in military professionalism by the military institution nor its individual elites. With these conditions, the military does not have an out-sized platform to interact with the American public, nor are there outstanding incentives for the military to raise significant disagreement over any domestic political issues.

\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{Table Two: Four Types of "Military Professionalism Worlds" Where Institutional Credibility and Political Polarization Vary} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    Influence of Variables & \textbf{Low Institutional Credibility} & \textbf{High Institutional Credibility}  \\ \hline
    \textbf{Low Political Polarization} & Negligible anticipated impacts. Traditional Professionalism is Expected by the Military, Civilian Leaders, and the American Public & Possible Impacts. There is a demand for military voice, but little substantive matter stemming from US domestic politics that can cause crucial disagreement. Little incentive to deviate from traditional concepts of loyalty and duty. \\ \hline
    \textbf{High Political Polarization} & Possible Impacts. Polarization raises the stakes of domestic politics, but military professionals are constrained from voicing their opinions and/or interfacing with the public due to low institutional credibility with the American public &  Highly Likely Impacts to Professionalism. Likely impacts. Military professionals have a "platform with the public" due to high institutional credibility, and high polarization raises the likelihood of disagreement between individual military elites and civilian leaders on a number of crucial topics.\\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing
The second possible set of conditions is when political polarization is low and institutional credibility is high. When these set of conditions exist, the military has a platform to interact with the public, as the public highly respects the military institution. The military enjoys this high level of trust and does not want to lose it. Further, because the degree of political polarization is low, there is little incentive for the military as an institution or military elites to use the trust they enjoy with the American public to adopt or denounce an issue associated with any partisan entity. The high institutional credibility increases the chance of the military interacting with the public in an unprofessional way, but the low political polarization removes the incentives of doing so.

@@

dataset <- read_csv(here("data", "MilEditorials10Aug.csv"))
dataset %>% print()
# Only read in the first 334 rows - is there a better way to get rid of N/A data?
# Replace NA with 0 in the data for variables War_Ops thru Fig_Nme
adjdataset <- dataset[1:334,] %>%
  mutate_at(
    .vars = vars(War_Ops:Fig_Nme),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

@@

@@

slimdata <- polaradjdataset[1:334,] %>% 
  view()

>>>>

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers since 1979. In doing so, this paper develops a theory which argues that increasing political polarization in the United States impacts the military as a profession by straining the individual military officer's concept of loyalty in important ways. Drawing on an original data set of more than 340 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold the concept of military professionalism as defined by historical military norms and standards. The paper concludes by arguing that the concept of "politicization of the military" --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the novel phenomenon of retired military officers appealing to one's concept of loyalty to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In many cities and communities across the nation, peaceful protests were followed in some cases by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America by authoring opinion-editorial publications. It is a question worthy of examination for many reasons, including that at its root, at least three factors are involved, each of which may shape civil-military relations: military professionalism, political polarization, and institutional credibility of the military. The main argument of this paper is that a weakening of the concept of military professionalism can clearly be seen in the way retired military officers have made arguments in the public square over the past four decades, that the main variable that can explain this variation is increasing political polarization within American Society. Further, this paper argues that the military institution is particularly susceptible to increasing political polarization because of the way the concept of loyalty is used to make arguments that contain a clear partisan or political leaning. These developments reveal the extent to which increasing political polarization is synonymous with a fractured domestic public, and highlight the extraordinary challenge that the US military will likely face to remain a  professional in an enduring polarized political climate.

This paper proceeds in four parts. First, the paper briefly reviews the concept of military professionalism in order to orient the reader to the central role this concept has played in the ethos of the modern American military. Second, the paper builds a theory that demonstrates how certain features of today's polarized political landscape challenge the concept of loyalty for the individual military professional. Third, the paper explores and analyzes a data set which consists of more than 340 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately shows that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to express clear partisan preferences when making their arguments. The analysis also shows that this trend has worsened as the American political climate has become increasingly polarized over the past several decades. Finally, the paper concludes by briefly exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting that will likely endure.

\singlespace
# Military Professionalism as a Foundational Trait
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is long and storied. As a concept, the idea of a professional military stems from the post-WW2 era, when the US military was locked in the Cold War struggle against the Soviet Union. Perhaps the strongest and certainly one of the original intellectual founders of a professional military is the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the character of American liberal society. Huntington further argued that the key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

@@

Although a professional and non-partisan military is the theoretical goal of the institution, the practical record and results remain far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory, in one form or another, on more than one occasion. Nonetheless, academic theorists, military tradition, and contemporary military leaders have long argued for the importance of the military remaining grounded in professional conduct that is both non-partisan and apolitical. At its core, it is this type of professional conduct which ultimately postures the military to earn the trust and confidence of the American people^[There are certainly other aspects that impact the military's ability to earn the public's trust, such as performing its military missions well, winning in combat, etc.].

To be clear, the concept of military professionalism does not impose a normative prohibition on the publication of an opinion piece, or the expression of other forms of public dialogue, by a military professional. In fact, one of the foundations of Huntington's concept of military professionalism is the fact that military leaders possess unique and specialized skills that civilians simply do not have. Moreover, multistar generals and admirals in the contemporary military possess in many cases upwards of 30 and sometimes 40 years of service, and it is entirely reasonable to conclude that these high-ranking officers in particular have been privy to decisions and experiences occurring at or near the highest levels of government. In other words, their experiences are not confined to only tactical matters. They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service, and much more.

It is reasonable, therefore, to conclude that a professional opinion piece written by a military officer would take on three qualities. First, a professionally-written opinion piece would address a relevant military topic consistent with the military expertise of the author. In the US context in particular, where the military conducts a variety of roles and performs a variety of missions, this would include a wide swath of topics, while at the same time, certain topics would be considered out of the realm of relevant military expertise, such as partisan domestic politics. Second, from the perspective of both expertise and institutional credibility, a professional opinion piece would be based on the officer's personal experiences and be written from an individual's perspective so as not to infer that an argument is being made on behalf of the military. Third, a professionally-written opinion piece would be free of of partisan sentiment or expression, whether advocacy or denouncement. If one of these three qualities -- addressing a relevant military topic, written from the perspective of an individual officer's expertise, free of partisan sentiment -- fails to be upheld, there is an increasing likelihood that the officer's words can be conveyed and/or received as unprofessional.

If and when a military officer violates standards of military professionalism when conveying an opinion to the American public, his or her words can cause great harm, either to American society, the military as an institution, or a combination of both. To understand how and why this is the case, we will now turn our attention to understanding the impacts on military professionalism caused by two changes that have developed within American Society in the past 50 years. The first is a gradual increase in the American public's opinion of the military as an institution, and the second, more significant change is the steady increase in political polarization within American society.

@@

\singlespace
# When Words Become Weaponized - A Theory of Political Polarization and Military Professionalism
\doublespace

@@

What mechanisms might be at work through increasing political polarization that ultimately impact military professionalism? To answer this important question, I rely on the three major characteristics of the contemporary polarized political landscape previously identified by scholars of American Politics. These included, first, a sharp increase in the number and scope of issues that have been subsumed by partisan debate and disagreement (the phenomenon referred to as "conflict extension" [@layman_party_2002]); second, the increase in the strength of negative feelings held by one partisan group towards those of the other side (the rise of "negative partisanship" [@abramowitz_great_2018]); and third, an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. Each of these three features of the politically polarized landscape strain the concept of military professionalism in important ways, and taken together, comprise a significant challenge to the individual military professional and thus, to the professional nature of the military as a whole. These challenges are summarized in Table 1, below.

@@

The first way in which the concept of military professionalism is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This happens because as "conflict extension" [@layman_party_2002] occurs, an increasing number of issues are brought into partisan battlegrounds, and thus, there is less politically-neutral territory. When polarization levels in America were not as high as they are today, it was more possible for individuals to either stay out of politics, avoid partisan squabbles, and otherwise to stay politically neutral. Today's polarizing environment seems to have the opposite impact: polarized politics seem to thrust issues and the unfolding political fight at the individual citizen, even when the individual is not deliberately seeking confrontation. There is no theoretical reason to suspect that this mechanism would not impact military leaders. Military leaders are citizens of the Republic, and though the institutional ethos of the military profession is to remain apolitical, such an ideal is difficult to uphold if less and less neutral political space exists.

@@

However, it is also precisely on this point that the military institution is perhaps most vulnerable to increasing political polarization. I contend that the rise of identity politics can and will be harmful if a military professional's identity and the loyalties associated with that identity directly conflict with his or her identity and loyalty as a professional soldier. As any soldier knows, and as civilians can reasonably understand,few traits are more sacred in the military than loyalty. Indeed, the virtue of loyalty is extolled in a host of codes and mantras that have played a central part of the American military since its inception, and loyalty is captured, either in word or in spirit, in a host of phrases and professional codes. In the middle of the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918]. Even Huntington spent a great deal talking about the importance of loyalty as a trait to the very essence of the military profession. He recognized that in order for civilian control to be effective, the military profession had to organize itself as a "hierarchy of obedience," which required "loyalty and obedience" to stand out as "the highest military virtues" [@huntington_soldier_1957, 73].

High levels of political polarization obfuscates the ideal of loyalty, however, and thus pose a significant risk to the professional nature of the military. As core issues that involve individual perceptions of moral right and wrong are at stake in an increasing number of partisan battles, individuals, to include military officers, are more likely to become attached to or otherwise invested in these issues. This is especially true when the justification for taking or not taking action invokes the concept of loyalty. As a result, highly polarized political climates increase the likelihood that two different military officers may adopt completely different opinions or stances on an issue, each believing that the other is not only morally wrong, but failing to act out of loyalty in accordance with a specific oath or moral code. A couple of recent cases illustrate how military officers, active and retired, are invoking the concept of loyalty to make an appeal to action that is, if not outright hostile to, certainly not helpful for, the ideal of military professionalism.

First, in June 2020, following President Trump's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@anonymous_authors_letter_2020]. The letter's clear implication is that Secretaries Esper and Pompeo have forsaken their oaths of office, the US Constitution, and the values that West Point stands for (the values of duty, honor, and country).

Then, in July 2020, several recent West Point graduates, who are now officers in the US Army, released an open letter addressed to the Superintendent, Commandant, Academic Dean, and Athletic Director of West Point. The letter's authors expressed concern that West Point had not done all it could do to be a "anti-racist" institution, and made a list of requests of the Academy's leadership, including "releasing a public statement acknowledging the existence of systemic racism at West Point, demonstrating commitment to anti-racism..., outlining anti-racist education..., and detailing plans to eradicate Confederate memorialization" [@askew_anti-racist_2020]. The authors further stated that, "we hope Academy leadership will agree that, by writing this proposal, we are espousing the same values that West Point seeks to instill in its graduates: leadership, duty, respect, selfless service, honor, integrity, and personal courage" [@askew_anti-racist_2020].

And in August, 2020, yet another open letter, this time written by a pair of retired Army Lieutenant Colonels, John Nagl and Paul Yingling, addressed the Chairman of the Joint Chiefs of Staff, General Mark Milley, invoked the concept of loyalty to frame a very disconcerting prospect: active duty military forces removing President Trump from office if, in January, he fails to accept the results of the election [@nagl__2020]. Nagl and Yingling assert that Milley has a choice to make. As the letter states to Milley, "the choice between these two options [_removing the President from office, or nothing_] lies with you...your duty is to give unambiguous orders directing U.S. military forces to support the Constitutional transfer of power. Should you remain silent, you will be complicit in a coup d’état [@nagl__2020,_emphasis mine].

Each of these three instances could deserve a much longer exposition and analysis, but the important thing to notice here as it relates to political polarization is how in each case, the concept of loyalty is used to justify taking action, but even more than that, to use the concept of loyalty to bring disdain and perhaps even shame on someone for failing to act in accordance with the letter's demands. This is remarkable in that in effect, political polarization has served to draw up battle lines for America's warriors, its venerated and historically apolitical institution.

@@

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table 2, below). The first is when political polarization and military credibility are both low. When these conditions hold, there is no anticipated drop in military professionalism by the military institution nor its individual elites. With these conditions, the military does not have an out-sized platform to interact with the American public, nor are there outstanding incentives for the military to raise significant disagreement over any domestic political issues. 
\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{Four Types of "Military Professionalism Worlds" Where Institutional Credibility and Political Polarization Vary} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    Influence of Variables & \textbf{Low Institutional Credibility} & \textbf{High Institutional Credibility}  \\ \hline
    \textbf{Low Political Polarization} & Negligible anticipated impacts. Traditional Professionalism is Expected by the Military, Civilian Leaders, and the American Public & Possible Impacts. There is a demand for military voice, but little substantive matter stemming from US domestic politics that can cause crucial disagreement. Little incentive to deviate from traditional concepts of loyalty and duty. \\ \hline
    \textbf{High Political Polarization} & Possible Impacts. Polarization raises the stakes of domestic politics, but military professionals are constrained from voicing their opinions and/or interfacing with the public due to low institutional credibility with the American public &  Highly Likely Impacts to Professionalism. Likely impacts. Military professionals have a "platform with the public" due to high institutional credibility, and high polarization raises the likelihood of disagreement between individual military elites and civilian leaders on a number of crucial topics.\\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing
The second possible set of conditions is when political polarization is low and institutional credibility is high. When these set of conditions exist, the military has a platform to interact with the public, as the public highly respects the military institution. The military enjoys this high level of trust and does not want to lose it. Further, because the degree of political polarization is low, there is little incentive for the military as an institution or military elites to use the trust they enjoy with the American public to adopt or denounce an issue associated with any partisan entity. The high institutional credibility increases the chance of the military interacting with the public in an unprofessional way, but the low political polarization removes the incentives of doing so.

@@

dataset <- read_csv(here("data", "MilEditorials10Aug.csv"))
dataset %>% print()
# Only read in the first 344 rows - is there a better way to get rid of N/A data?
# Replace NA with 0 in the data for variables War_Ops thru Fig_Nme
adjdataset <- dataset[1:344,] %>%
  mutate_at(
    .vars = vars(War_Ops:Fig_Nme),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

@@

# Create a new variable called Yrcount, which is a count of the yearly number of published op-eds

adjdataset <- adjdataset %>% 
  group_by(PubYr) %>% 
  mutate(Yrcount=n()) 
```

Ultimately, I aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened dramatically. To answer this question, I assemble and examine a data set that consists of opinion-editorial publications authored by retired military officers in major US newspapers. My methodological approach involves two fundamental steps. First, I determine what topics retired military officers have been writing about over time. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed (apolitical in nature, non-partisan, relying on relevant military expertise, etc.). In essence, I want to answer the questions, what are these retired military officers writing about, and then, using the normative approach to military professionalism just discussed, are these op-ed's deviating from the concept of military professionalism, and how?

To assemble the data, I use the Factiva and Nexus-Uni databases, conducting a search that focuses keywords in the byline (author field) such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth. I confine my search to the largest US newspapers that regularly run opinion-editorial pieces, which includes  _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. The time period for the data includes the period January 1, 1979 - 30 August, 2020, the main limitation being that not all of the Newspaper sources are available for previous date periods. I also open my search to include both print and online offerings in order to ensure that I am capturing any opinion pieces published digitally. The data only include publications classified as either editorials or commentaries/opinions by the databases. Thus, other types of pieces that may still be authored by retired military officers, such as letters to the editor and/or published interviews, are not included.

The data set contains 344 observations, penned by 185 different authors or author teams.^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] Several authors penned multiple pieces, including Retired Army Lieutenant General (23 publications) and Retired Air Force General Michael Hayden (20 publications). Ten publications were written anonymously, with at best general recognition by the authors to their previous military service and/or rank. From a macro perspective, consider for a moment all that the date range includes with respect to the US military. Indeed, numerous and substantive events, debates, and changes have occurred in this time period for retired military officers to weigh in on. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, as well as a host of smaller but important military operations in Grenada, Panama, Somalia, and Haiti. The military has also undergone significant social and institutional changes in the time period examined. In the late 1970s, the military had ended the draft only years before following the conclusion of the Vietnam War. The debates over "Don't Ask, Don't Tell" and policies regarding women in the combat arms are also included in the data, as are numerous discussions of military budgets, weapons systems, and alliances. It is also interesting to note, as shown in Figure Two, below, that retired military officers have increasingly written and successfully published opinion-editorial pieces in the sources examined over the past forty years.

```{r figure two, fig.height=4, fig.width=5, fig.align="center", warning=FALSE, include=TRUE, echo=FALSE}

# construct a scatter plot with line and smoothing

ggplot(data=adjdataset, aes(x=PubYr, y=Yrcount)) + geom_point() + geom_line() + geom_smooth(method="loess") + labs(
  x="Year", 
  y="count", 
  title="Figure Two: Number of Op-Eds Published by Retired\n Military Officers (WSJ, NYT, WaPo, USA Today, and LA Times",
  subtitle= "1979-2020",
  caption="Note: As of 31 August 2020"
  ) + scale_x_continuous(limits=c(1979, 2020)) + scale_y_continuous(limits=c(-7, 28)) + 
geom_vline(xintercept = 1981, size=.2, colour="tomato") + 
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=-5), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=-5), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=-5), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=-5), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=-5), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=-5), colour="tomato", angle=90, vjust=1, text=element_text(size=9))

@@

slimdata <- polaradjdataset[1:344,] %>% 
  view()
